Whole-genome pairwise alignment using the hierarchical wavefront algorithm

### 2022-04-01 manuscript/main.tex
abstract

\abstract{
\textbf{Motivation:}
Pairwise alignment is a critical step in studying whole-genome assemblies and their relationships in an unbiased way.
The time and memory required to compute the pairwise alignments increase quadratically with sequence length and nucleotide divergence.
This makes it impractical to align very long sequences without applying heuristic approaches to first determine possible synthenic regions in the sequences.
Nevertheless, with the advances in sequencing technologies, new whole-genome assemblies are produced at a high rate, pressing for the development of tools able to align sequences of the order of tens of megabases long. \\
\textbf{Results:}
Here we present wfmash, a new gap-affine pairwise aligner designed to align DNA sequences at a whole-chromosome scale.
wfmash applies a hierarchical implementation of the wavefront alignment algorithm to guide the alignment of very long sequences.
It scales efficiently to large eukaryotic chromosomes, allowing users to perform pairwise alignment of thousands of large genomes using little time and memory. \\
\textbf{Availability:}
wfmash is published as free software under the MIT open source license.
Source code and documentation are available at \url{https://github.com/ekg/wfmash}.
wfmash can be installed via Bioconda \url{https://bioconda.github.io/recipes/wfmash/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/wfmash.scm}. \\
\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
}

>>>>

\abstract{
\textbf{Motivation:}
Pairwise alignment is essential to understanding DNA sequence variation.
The time and memory required to compute pairwise alignments increases quadratically with sequence length and nucleotide divergence.
Advances in sequencing technology have resulted in the rapid production of new whole-genome sequences.
This leads to a new problem of efficiently comparing many whole genomes, a costly task for methods designed to align shorter (100bp--100kbp) sequence reads to reference genomes. \\
%This makes it impractical to align very long sequences without applying heuristic approaches to first determine possible synthenic regions in the sequences.
%Nevertheless, with the advances in sequencing technologies, new whole-genome assemblies are produced at a high rate, pressing for the development of tools able to align sequences of the order of tens of megabases long. \\
\textbf{Results:}
Here we present \textit{wfmash}, a sequence aligner specialized for large collections of very long and highly divergent genome sequences.
It first finds large syntenic homologies between genomes using locality sensitive hashing, approximate match chaining, and logarithmic-time filtering of these mappings.
%Two parameters define these mappings: a seed segment length and minimum average nucleotide identity between matched segments, allowing for intuitive experimental setup.
The resulting mappings can cover entire chromosomes and contain megabase-scale SVs, which presents a significant challenge for direct alignment.
We address this with an extension of the wavefront algorithm (WFA) that matches segments of sequence rather than single characters.
Our approach achieves a many thousand-fold reduction in time and space complexity relative to direct alignment with WFA.
We demonstrate that this comes at little to no cost in terms of variant detection accuracy relative to established sequence aligners. \\
%\textit{wfmash} allows us to scale to all-to-all alignment of hundreds of vertebrate \\
%A key feature of \textit{wfmash} is that users can intuitively shape the scope of an alignment experiment using two parameters: a mapping segment length and a minimum average nucleotide identity between segments. \\
\textbf{Availability:}
wfmash is published as free software under the MIT open source license.
Source code and documentation are available at \url{https://github.com/ekg/wfmash}.
wfmash can be installed via Bioconda \url{https://bioconda.github.io/recipes/wfmash/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/wfmash.scm}. \\
\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
}
